An Adolescent Mini-Neptune in the Kepler Field

### 2021-07-08 paper/ms.tex
add get_flare_pair_time_distribution
update ms
flare analysis

\begin{abstract}
  Kepler\,1627A is a G8V star previously found to host a
  $3.6\,R_\oplus$ mini-Neptune on a 7.2\,day orbit.  The star was
  selected for observation by the Kepler satellite because it is
  nearby ($d\approx 333\,{\rm pc}$) and resembles the Sun.  Combining
  data from Gaia with rotation periods from TESS, we show that
  Kepler\,1627 is a member of the $35\pm XX$\,Myr old $\delta$~Lyr
  cluster.  Kepler\,1627Ab is therefore the youngest planet with a
  precise age found so far by the main Kepler mission.
  Analyzing the photometry, we find evidence that starspot crossings
  distort the average transit shape, and induce transit timing
  variations that suggest the planet is on a prograde orbit.  The
  possibility that star-planet interactions could induce
  flares in the system is also explored.  Future spectroscopic studies
  of \pn\ may help clarify how the orbit and atmosphere of this
  mini-Neptune evolve.  More generally, the combination of Gaia and TESS
  offers an opportunity to significantly expand the census of
  age-dated planets; we provide a literature concatenation of young
  stars to support this effort.
\end{abstract}

@@

% Could the planet be on a high obliquity orbit?
% * how do the spectroscopic vsini and the measured v_rot=2πRstar/Prot compare?
%   * Rstar=0.91+-0.05, 2.606 days gives vsini=17.6km/s (+/-0.97km/s)
% * spectroscopic vsini is 18.8 +/- 4.7km/s from Keck HIRES... "20km/s" from
%   TRES.

@@

The 1-minute cadence Kepler observations span 98 days, and cover 24
flares exceeding $0.5\%$ in relative flux.  These 24 flares spanned a
total of 6.5 hours ($\sim$15 minutes per flare).  The coincidence is
that despite the low flare duty cycle, one orbital period after the
brightest flare, a second flare followed.  This and a similar
event are shown in Figure~\ref{fig:flarezoom}.  The timing error is
good to a $\approx0.2\%$ difference from the orbital period, which
seems {\it a priori} unlikely.  If we consider flares falling within
2\% of the planet's orbital period after a previous flare, then 4 of
the 24 flare events have candidate ``successors''.

A brief note on how we cleaned the light curve, and identified the
flares.
For cleaning, we performed the following iterative detrending procedure.
\begin{itemize}
  \item Step 1: Build a two-term mixed SHOTerm GP model with
    quasi-periodic kernels at Prot and 0.5$\times$Prot. Fit the model
    to the time, flux, and flux uncertainty.
  \item Step 2: Select points more than twice the median absolute
    deviation from the residual, and exclude them from the light
    curve.  Repeat Step 1.
  \item Step 3: On the residual from Step 2, identify all flares,
    requiring them to be at least 20 cadences apart, at least 7 median
    absolute deviations above the median baseline, and lasting at least
    2 cadences in duration.  Build the mask spanning these times, from
    5 minutes before each flare begins to 2.5 minutes after the final
    flare cadence.  Repeat Step 1 a final time.
\end{itemize}

The flares were identified and fitted using ALTAIPONY (CITE CITE).
The fitted model is that of CITE XXX, which parametrizes the flare
with a start time, a lag time, and an amplitude (CHECK).
Figure~\ref{fig:flarephase} shows the resulting flares, amplitudes,
and phases.

We considered two hypotheses: {\it i)} the flare arrival times are
distributed as a Poisson process, and {\it ii)} they are better
explained as a Poisson process with a periodic mixture component.

TODO: explore!  Fit the flare inter-arrival times...
[even the 2-separated ones?]

>>>>

\begin{abstract}
  Kepler\,1627A is a G8V star previously found to host a
  $3.6\,R_\oplus$ mini-Neptune on a 7.2\,day orbit.  The star was
  selected for observation by the Kepler satellite because it is
  nearby ($d\approx 333\,{\rm pc}$) and it resembles the Sun.  Combining
  data from Gaia with rotation periods from TESS, we show that
  Kepler\,1627 is a member of the $35\pm XX$\,Myr old $\delta$~Lyr
  cluster.  Kepler\,1627Ab is therefore the youngest planet with a
  precise age found to date by the main Kepler mission.
  Analyzing the photometry, we find evidence that starspot crossings
  distort the average transit shape, and induce transit timing
  variations that suggest the planet is on a prograde orbit.  The 
  flare arrival times may show hints of being
  non-Poissonian, but the statistical evidence is not overwhelming ($<3$-$\sigma$).  In the future,
  spectroscopic studies of \pn\ may help clarify the evolution of this
  mini-Neptune's orbit and atmosphere.  More broadly, the
  combination of Gaia and TESS data offers an opportunity to
  expand the census of age-dated planets; we provide a literature
  concatenation of young stars to support this effort.
\end{abstract}

@@

@@

\begin{figure*}[t]
	\begin{center}
		\leavevmode
		\includegraphics[width=1.0\textwidth]{f13.pdf}
	\end{center}
	\vspace{-0.7cm}
	\caption{
		{\bf Statistics of inter-flare arrival times}.  
    24 flares were recorded with amplitudes exceeding 0.5\% over the
    97.7 days of short cadence observations.  The histogram of the
    time intervals between every possible pair of flares is shown in
    black.  Some plausibly important timescales for star-planet
    interactions, namely the planetary orbital period and synodic
    period (the orbital period as seen from the rotating stellar frame) are
    shown along with their linear combinations.  Monte Carlo
    draws from a Poisson distribution are shown
    with the gray bands.  While peaks in the observed distribution do
    coincide with the locations of these ``special periods'', the
    statistical evidence for a non-Poissonian process driving the
    flares does not reach the $5$-$\sigma$ threshold.
		\label{fig:flarestats}
	}
\end{figure*}

The short cadence (1-minute) Kepler observations span 97.7 days.
The light curve shows a significant number of flares
(Figure~\ref{fig:flarezoom}).
To quantify clean the light curve and identify the flares, we
performed the following procedure.
For cleaning, we performed the following iterative detrending procedure.
\begin{itemize}
  \item Step 1: Build a two-term mixed SHOTerm GP model with
    quasi-periodic kernels at Prot and 0.5$\times$Prot. Fit the model
    to the time, flux, and flux uncertainty.
  \item Step 2: Select points more than twice the median absolute
    deviation from the residual, and exclude them from the light
    curve.  Repeat Step 1.
  \item Step 3: On the residual from Step 2, identify all flares,
    requiring them to be at least 20 cadences apart, at least 7 median
    absolute deviations above the median baseline, and lasting at least
    2 cadences in duration.  Build the mask spanning these times, from
    5 minutes before each flare begins to 2.5 minutes after the final
    flare cadence.  Repeat Step 1 a final time.
\end{itemize}
The flares were then identified and fitted using ALTAIPONY (CITE CITE).
The fitted model is that of CITE XXX, which parametrizes the flare
with a start time, a lag time, and an amplitude (CHECK).
Figure~\ref{fig:flarephase} shows the resulting flares, amplitudes,
and phases.

There were $N_{\rm f}=24$ flares exceeding $0.5\%$ in relative flux during the
short-cadence observations.  These 24 flares spanned a total of 6.5
hours ($\sim$15 minutes per flare).  For better or worse, we noticed a
coincidence in the flare arrival times.  The coincidence is that
despite the low flare duty cycle, one orbital period after the
brightest flare, a second flare followed.  This and a similar event
are shown in Figure~\ref{fig:flarezoom}.  The timing error is good to
a $\approx0.2\%$ difference from the orbital period, which seems {\it
a priori} unlikely.  If we consider flares falling within 2\% of the
planet's orbital period after a previous flare, then 4 of the 24 flare
events have candidate ``successors''.

As with any coincidence, when one does not have a firm prediction, it
is difficult to assess the statistical significance of a surprise.
Since our surprise was specifically at the inter-arrival time of
certain flares coinciding with special time intervals, we performed
the following analysis.  First, we considered all unordered pairs of
flares.  For $N$ flares there are ${n \choose 2}$ such pairs (for our
case, 276 pairs).  We then compared the distribution of the pair
separations against that from a Poisson distribution.  Specifically,
we drew $N_{\rm f}=24$ samples from a Poisson distribution with
$\lambda = \Delta t / N_{\rm f}$, for $\Delta t=97.7\,{\rm days}$ the
duration of the observations, and repeated the draw $10^3$ times with
unique random seeds.

Figure~\ref{fig:flarestats} shows the results.  The vertical lines in
the figure show the planetary orbital period, the synodic period
$P_{\rm syn} = (P_{\rm rot}^{-1} - P_{\rm orb}^{-1})^{-1}$, and linear
combinations thereof.  Note that the tidal period (half the synodic
period) is not shown.  The bins are logarithmically spaced to give 100
bins between the minimum and maximum ordinate values.  The gray bands
express the range of values observed from the Poissonian draws.  While
it does seem like a rather odd coincidence for peaks in the observed
flare arrival time distribution to coincide with the locations of
these ``special intervals'', the statistical evidence for a
non-Poissonian process driving the flares does not seem particularly
strong.  More quantitatively, the peaks observed at the orbital and
synodic periods are within the $\pm 2$-$\sigma$ range of a
Poissonian process, and those at $P_{\rm orb}+P_{\rm syn}$ and $P_{\rm
orb}+2P_{\rm syn}$ are only slightly above this range. 
With that said,
future analyses of these data by investigators more versed in the
topic than ourselves could very well yield new insights.
